Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-06-30 manuscript/manuscript_file.Rmd
Minor updates to manuscript.

-   Table and figure captions should specify if displaying sample vs population estimates. "posterior predictive distribution"
-   Note that right tail is truncated in figure 2.

@@

The extent of N~2~O undersaturation in CONUS waterbodies is surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long standing paradigm that surface waters are a source of N~2~O to the atmosphere (@mosier_etal_1998) and joins other recent reports that lentic waterbodies can function as an N~2~O sink. @Webb_etal_2019 found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months and @Soued_etal_2016 report that 40% of rivers, lakes, and ponds sampled during the summer in a boreal region of Quebec, Canada were functioning as N~2~O sinks **[29% of lakes in Finland (kortelainen et al 2020) but kortelainen and Soued are mostly pristine lakes]**. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months. This pattern holds across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use across the continental U.S..

@@

# create flextable
flextable(table1_dat) %>%
  set_header_labels('waterbody surface area' = "waterbody surface area\n(hectares)",
                    'N2O undersaturation extent' = "N2O undersaturation extent\n(% waterbodies)",
                    'N2O emission rate' = "N2O emission rate\n(mg N2O m-2 d-1)",
                    'N2O flux' = "N2O flux\n(Mg N2O y-1)") %>%
  align(j=c(2:6), align = "center", part = "all") %>% # center align columns 2:4
  valign(valign = "top", part = "header") %>% # top align column names
  # caption doesn't show in .Rmd viewer, but renders correctly in .docx
  set_caption(
    caption = as_paragraph( # allows for formatting
      as_chunk("Table 1. Mean and 95% confidence interval of N2O undersaturation extent, saturation ratio, emission rate, and flux at the national and ecoregional scales", 
               props = fp_text_default(bold = TRUE) # here is where we specify formatting (font, size, bold, etc)
      )
    ),
    align_with_table = FALSE) %>% # table is center aligned, ignore that and apply default left align to caption
  autofit() %>% # convenient column spacing
  fontsize(size = 8, part = "all") # doesn't apply to caption when as_paragraph is invoked.

@@

Dissolved N~2~O concentration increased with NO~3~ concentration **[Fig. 3]**, likely because NO~3~ concentration partially controls denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ [@knowles_1982, @richardson_etal_2009]. The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column. Incomplete denitrification often occurs when NO~3~ is relatively abundant [@Beaulieu_etal_2015, @Beaulieu_etal_2014], likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification [@koike_etal_1975, @richardson_etal_2009]. When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~, thereby functioning as an N~2~O sink.  N~2~O consumption via denitrification likely explains the N~2~O undersaturation observed at low NO~3~ concentrations, whereas denitrification was likely a source of N~2~O at higher NO~3~ concentrations.

It is also possible that the correlation between N~2~O and NO~3~ reflects the simultaneous production of these solutes via nitrification. Nitrification, an aerobic process which can produce N~2~O as a byproduct, is the microbial oxidation of ammonium (NH~4~^+^) to NO~3~ **[citation]**. Nitrification might be expected to be more prevalent in the predominantly well oxygenated surface waters where the N~2~O samples were collected than denitrification, which is an anaeorbic process. However, denitrification can occur in anoxic microsites within particulate matter suspended in an aerobic water column **[citation]** and can be transported from aoxic sediments to the water surface via advection and diffusion [@delsontro_etal_2018]. Thus the N~2~O - NO3 relationship may be reflective of some combination of N~2~O production/consumption via denitrification and N~2~O production via nitrification.

![Figure 3.  A) Relationship between N~2~O saturation and nitrate (NO~3~) concentration, dependent upon surface water temperature.  The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and < 4.09, 3) $\geq$ 4.09 and < 30.20, 4) $\geq$ 30.20 and < 223.13, and 5) $\geq$ 223.13. B) Relationship between N~2~O saturation and surface water temperature, dependent upon NO~3~ category.  Horizontal dashed lines in both panels reflects an N~2~O saturation value of 1.](manuscript_figures/sat_cond_surftemp.tiff)

@@

While NO~3~ and water temperature were the primary drivers of N~2~O production or consumption rates, dissolved N~2~O concentration was also related to waterbody size.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases with increasing waterbody size class **[Fig. 4A]**. This pattern holds regardless of whether N~2~O is supersaturated or undersatured **[Fig. 4B, C]**.  Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In lakes with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column (@carpenter_1983). Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface (@Beaulieu_etal_2015, @Beaulieu_etal_2014).

@@

![Figure 4.  A) Relationship between $\Delta$ N~2~O (difference between observed and equilibrium N~2~O concentration) and waterbody surface areas.  B,C) relationship between N~2~O saturation and waterbody size for the Coastal Plains and Northern Plains ecoregions.  The vertical blue line references a N~2~O saturation value of 1.0.  The horizontal red arrows highlight $\Delta$ N~2~O for the smallest lake size class.](manuscript_figures/deltaN2ObySize.tiff)

@@

The model indicates that NO~3~, water temperature, and surface area interact to determine the N~2~O saturation ratio. These interactions yield a broad range of N~2~O saturation values across the population of CONUS waterbodies, but also enable the identifcation of broad scale patterns. First, small waterbodies with low NO3 concentration tend to be undersaturated, whereas small waterbodies with high NO3 tend to be highly supersaturated. Large waterbodies, regardless of NO~3~ status, tend to have N~2~O saturation ratios closer to equilibrium. Therefore intra and inter regional patterns in N~2~O saturation are determined by the joint distributions of waterbody size, NO~3~ availability, and water temperature.

![Figure 5. Partial dependence plots illustrating the relationship between N~2~O saturation and A) nitrate (NO~3~ availability dependent upon lake size, and B) lake size dependent upon NO~3~ availability. The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and < 4.09, 3) $\geq$ 4.09 and < 30.20, 4) $\geq$ 30.20 and < 223.13, and 5) $\geq$ 223.13.](manuscript_figures/sat_cond_area.tiff)

**Not sure where I am going with this paragraph. Maybe scratch it. Mostly included as a placeholder for these references** While previous surveys have documented relationships between lake size and dissolved methane concentration (Pighini et al. 2018, Juutinen et al 2009, Borges et al. 2011), another greenhouse gas, to the best of our knowledge, this is the first study to document a relationship between lake size and $\Delta$ N~2~O. Perhaps because previous surveys have included little variation in waterbody size (@webb_etal_2019, @kortelainen_etal_2020) or water quality (@soued_etal_2016).

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was `r national_stats$post_med_en2o` (95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N2O m^-2^ d^-1^ (Table 1). This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N2O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N2O m^-2^ d^-1^) (SI Figure 1). At the ecoregional scale, only the Northern Appalachians and Northern Plains ecoregions had mean emission rates less than 0.  These ecoregions also had the lowest mean NO~3~ availability of all nine ecoregions **[need table or figure.  See ecoreg_stats$post_med_no3 variable.  maybe a SI table with summaries of all data from all_predictions?]**.

@@

At the national scale, we estimate that CONUS waterbodies emit [problem knitting inline code? table1_dat %>% filter(`spatial domain` == "national") %>% select(`N2O flux`) %>% pull()] 1049 (358 - 1581) Mg N~2~O year^{-1}, assuming that the emission rates reported here hold for an entire calendar year. This is likely a conservative estimate because our data is based on summer conditions, whereas several studies have reported maximum N~2~O emission rates during the winter (citation, see climate benefits paper) and Fall (harsha n2o paper, LD n2o paper) in lakes and reservoirs.  While we are unaware of independent estimates of N~2~O emissions from CONUS lakes and reservoirs, the Intergovernmental Panel on Climate Change estimates that CONUS drainage networks, inclusive of zero order streams through major river mouths, receiving run off from croplands and grasslands emit `r ipcc.n2o.surface` (22,767) Mg N~2~O year^{-1} (@epa_2023).  Emissions from lakes and reservoirs represent only `r round((national_stats %>% select(post_med_fn2o) %>% pull() / ipcc.n2o.surface) * 100, digits = 1)` % of total surface water emissions reported by the IPCC (@epa_2023), with the balance emanating from streams and rivers.  Emissions from CONUS lakes and reservoirs are also small when compared to other anthropogenic N~2~O sources including wastewater treatment (88,00  Mg N~2~O year^{-1}), stationary combustion (84,000 Mg N~2~O year^{-1}), and manure management (66,000  Mg N~2~O year^{-1}).

![Figure 6: N~2~O flux by ecoregion and lake size category](manuscript_figures/n2oFluxBySizeEco.tiff)

@@

The smallest lake size classes exhibited both the highest and lowest levels of N~2~O saturation, depending on NO~3~ availability.  It is unclear whether this variability is related to differences in NO~3~ loading or temporal variation.  For example, reservoirs have been shown to function as N~2~O sources in early summer when NO~3~ is replete, but switch to N~2~O sinks in later summer when the NO~3~ supply has been exhausted (@beaulieu_etal_2014).  A better understanding of the temporal evolution of N~2~O production and consumption in small waterbodies is an important research need to determine the extent to which these systems can remove excess nitrogen from surface waters and simulatenously mitigate N~2~O emissions.

![SI Figure 1. Distribution of N~2~O emission rates by ecoregion.  Data are inferences made to the population of CONUS waterbodies.](manuscript_figures/SIfigure1.tiff)

>>>>

- include a better description of full range of lake size: sample and population.  also add NO3 to better support emission rate statement.
- incorporate in Methods section how/why size is incorporated as categorical variable and continuos variable.
- maybe discuss upper midwest.  Lots of lake area, but largely at equilibrium, therefore very small fluxes.  Large area doesn't necessarily mean big flux.
-  last sentence of emission rates.  we should look at summary stats of observed NO3 rather than predicted 
-  Mike will take a stab at big SI table.

@@

The extent of N~2~O undersaturation in CONUS waterbodies is surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long standing paradigm that surface waters are a source of N~2~O to the atmosphere (@Mosier_etal_1998) and joins other recent reports that lentic waterbodies can function as an N~2~O sink. @Webb_etal_2019 found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months and @Soued_etal_2016 report that 40% of rivers, lakes, and ponds sampled during the summer in a boreal region of Quebec, Canada were functioning as N~2~O sinks **[29% of lakes in Finland (kortelainen et al 2020) but kortelainen and Soued are mostly pristine lakes]**. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months. This pattern holds across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use across the continental U.S..

@@

# create flextable
flextable(table1_dat) %>%
  set_header_labels('waterbody surface area' = "total waterbody surface area\n(hectares)",
                    'N2O undersaturation extent' = "N2O undersaturation extent\n(% waterbodies)",
                    'N2O emission rate' = "N2O emission rate\n(mg N2O m-2 d-1)",
                    'N2O flux' = "N2O flux\n(metric ton N2O y-1)") %>% # 1 megagram (Mg) = 1 1000 kg = 1 metric ton
  align(j=c(2:6), align = "center", part = "all") %>% # center align columns 2:4
  valign(valign = "top", part = "header") %>% # top align column names
  # caption doesn't show in .Rmd viewer, but renders correctly in .docx
  set_caption(
    caption = as_paragraph( # allows for formatting
      as_chunk("Table 1. Mean and 95% confidence interval of N2O undersaturation extent, saturation ratio, emission rate, and flux at the national and ecoregional scales.", 
               props = fp_text_default(bold = TRUE) # here is where we specify formatting (font, size, bold, etc)
      )
    ),
    align_with_table = FALSE) %>% # table is center aligned, ignore that and apply default left align to caption

  footnote(i = 1, j = 3:6,
           ref_symbols = "a",
           value = as_paragraph("Mean and 95% credible interval of 500 realizations of posterior predictive distribution"),
           part = "header"
  ) %>%
    footnote(i = 1, j = 3,
           ref_symbols = ", b",
           value = as_paragraph("Percent of population with N2O saturation ratio < 1."),
           part = "header",
           inline = TRUE
  ) %>%
  footnote(i = 1, j = 5:6,
           ref_symbols = ", c",
           value = as_paragraph("Negative values indicate N2O uptake"),
           part = "header",
           inline = TRUE
  ) %>%
  fontsize(size = 8, part = "all") %>% # doesn't apply to caption when as_paragraph is invoked.
  fontsize(size = 7, part = "footer") %>% #smaller font for footnote
    autofit() %>% # convenient column spacing
  fit_to_width(ft_1, max_width = 11) # scale to fill 8x11 paper, landscape

@@

Dissolved N~2~O concentration increased with NO~3~ concentration **[Fig. 3]**, likely because NO~3~ concentration partially controls denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ [@Knowles_1982, @Richardson_etal_2009]. The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column. Incomplete denitrification often occurs when NO~3~ is relatively abundant [@Beaulieu_etal_2015, @Beaulieu_etal_2014], likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification [@Koike_etal_1975, @Richardson_etal_2009]. When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~, thereby functioning as an N~2~O sink.  N~2~O consumption via denitrification likely explains the N~2~O undersaturation observed at low NO~3~ concentrations, whereas denitrification was likely a source of N~2~O at higher NO~3~ concentrations.

It is also possible that the correlation between N~2~O and NO~3~ reflects the simultaneous production of these solutes via nitrification. Nitrification, an aerobic process which can produce N~2~O as a byproduct, is the microbial oxidation of ammonium (NH~4~^+^) to NO~3~ **[citation]**. Nitrification might be expected to be more prevalent in the predominantly well oxygenated surface waters where the N~2~O samples were collected than denitrification, which is an anaeorbic process. However, denitrification can occur in anoxic microsites within particulate matter suspended in an aerobic water column **[citation]** and can be transported from aoxic sediments to the water surface via advection and diffusion [@DelSontro_etal_2018]. Thus the N~2~O - NO3 relationship may be reflective of some combination of N~2~O production/consumption via denitrification and N~2~O production via nitrification.

![Figure 3.  Conditional effects plot illustrating A) the relationship between N~2~O saturation and nitrate (NO~3~) concentration, dependent upon surface water temperature.  The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and < 4.09, 3) $\geq$ 4.09 and < 30.20, 4) $\geq$ 30.20 and < 223.13, and 5) $\geq$ 223.13, and B) relationship between N~2~O saturation and surface water temperature, dependent upon NO~3~ category.  Horizontal dashed lines in both panels reflects an N~2~O saturation value of 1.](manuscript_figures/sat_cond_surftemp.tiff)

@@

While NO~3~ and water temperature were the primary drivers of N~2~O production or consumption rates, dissolved N~2~O concentration was also related to waterbody size.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases with increasing waterbody size class **[Fig. 4A]**. This pattern holds regardless of whether N~2~O is supersaturated or undersatured **[Fig. 4B, C]**.  Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In lakes with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column (@Carpenter_1983). Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface (@Beaulieu_etal_2015, @Beaulieu_etal_2014).

@@

![Figure 4.  A) Relationship between $\Delta$ N~2~O (difference between observed and equilibrium N~2~O concentration) and waterbody surface areas.  B,C) relationship between N~2~O saturation and waterbody size for the Coastal Plains and Northern Plains ecoregions.  The vertical blue line references a N~2~O saturation value of 1.0.  The horizontal red arrows highlight $\Delta$ N~2~O for the smallest lake size class.  Mean (points) and 95% credible interview (horizontal bars) are derived from 500 realizations of the poterior predictive distribution. ](manuscript_figures/deltaN2ObySize.tiff)

@@

The model indicates that NO~3~, water temperature, and surface area interact to determine the N~2~O saturation ratio. These interactions yield a broad range of N~2~O saturation values across the population of CONUS waterbodies, but also enable the identifcation of broad scale patterns. First, small waterbodies with low NO3 concentration tend to be undersaturated, whereas small waterbodies with high NO3 tend to be highly supersaturated. Large waterbodies, regardless of NO~3~ status, tend to have N~2~O saturation ratios closer to equilibrium. Therefore intra and inter regional patterns in N~2~O saturation can be well predicted by the joint distributions of waterbody size, NO~3~ availability, and water temperature.

![Figure 5. Conditional effects plots illustrating the relationship between N~2~O saturation and A) nitrate (NO~3~ availability dependent upon lake size, and B) lake size dependent upon NO~3~ availability. The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and < 4.09, 3) $\geq$ 4.09 and < 30.20, 4) $\geq$ 30.20 and < 223.13, and 5) $\geq$ 223.13.](manuscript_figures/sat_cond_area.tiff)

**Not sure where I am going with this paragraph. Maybe scratch it. Mostly included as a placeholder for these references** While previous surveys have documented relationships between lake size and dissolved methane concentration (Pighini et al. 2018, Juutinen et al 2009, Borges et al. 2011), another greenhouse gas, to the best of our knowledge, this is the first study to document a relationship between lake size and $\Delta$ N~2~O. Perhaps because previous surveys have included little variation in waterbody size (@Webb_etal_2019, @Kortelainen_etal_2020) or water quality (@Soued_etal_2016).

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was `r national_stats$post_med_en2o` (95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^ (Table 1). This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) (SI Figure 1). At the ecoregional scale, only the Northern Appalachians and Northern Plains ecoregions had mean emission rates less than 0.  These ecoregions also had the lowest mean NO~3~ availability of all nine ecoregions **[need table or figure.  See ecoreg_stats$post_med_no3 variable.  maybe a SI table with summaries of all data from all_predictions?]**.

@@

At the national scale, we estimate that CONUS waterbodies emit [problem knitting inline code? table1_dat %>% filter(`spatial domain` == "national") %>% select(`N2O flux`) %>% pull()] 1049 (358 - 1581) metric tons N~2~O year^-1^, assuming that the emission rates reported here hold for an entire calendar year. While we are unaware of independent estimates of N~2~O emissions from CONUS lakes and reservoirs, the Intergovernmental Panel on Climate Change estimates that CONUS drainage networks, inclusive of zero order streams through major river mouths, receiving run off from croplands and grasslands emit `r format(ipcc.n2o.surface, scientific = FALSE, big.mark = ",")` metric tons N~2~O year^-1^ (@epa_2023).  Emissions from lakes and reservoirs represent only `r round((national_stats %>% select(post_med_fn2o) %>% pull() / ipcc.n2o.surface) * 100, digits = 1)` % of total surface water emissions reported by the IPCC (@epa_2023), with the balance emanating from streams and rivers.  Emissions from CONUS lakes and reservoirs are also small when compared to other anthropogenic N~2~O sources including wastewater treatment (88,00 metric tons N~2~O year^-1^), stationary combustion (84,000 metric tons N~2~O year^-1^), and manure management (66,000 metric tons N~2~O year^-1^).

Our annual flux estimates assume that emission rate estimates based on samples collected during the summer reflect annual conditions.  The literature contains strong evidence of seasonality in N~2~O emission rates, but the patterns differ among studies.  Some systems have maximum N~2~O emission rates during the winter (@Musenze_etal_2014, @Kortelainen_etal_2020), others in the Fall (@Beaulieu_etal_2014), and yet others during the summer (@Beaulieu_etal_2010).  While our assumption of constant emissions across the year adds uncertainty to the annual flux estimate, it is not clear that the estimate is biased high or low.

![Figure 6: N~2~O flux by ecoregion and lake size category.  Mean (points) and 95% credible interval (horizontal bars) reflect the distribution of 500 realizations of the posterior predictive distribution.](manuscript_figures/n2oFluxBySizeEco.tiff)

@@

The smallest lake size classes exhibited both the highest and lowest levels of N~2~O saturation, depending on NO~3~ availability.  It is unclear whether this variability is related to differences in NO~3~ loading or temporal variation.  For example, reservoirs have been shown to function as N~2~O sources in early summer when NO~3~ is replete, but switch to N~2~O sinks in later summer when the NO~3~ supply has been exhausted (@Beaulieu_etal_2014).  A better understanding of the temporal evolution of N~2~O production and consumption in small waterbodies is an important research need to determine the extent to which these systems can remove excess nitrogen from surface waters and simultaneously mitigate N~2~O emissions.

![SI Figure 1. Distribution of N~2~O emission rates by ecoregion. Each blue line (n=500) represents the distribution of values from a single realization of the posterior prediction.](manuscript_figures/SIfigure1.tiff)
